---
title: "eda"
author: "Leonor Rui"
date: "2024-10-03"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(haven)
```

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10, 
  month = floor_date(date, unit = "month")) |>
  select(name, id, everything())
```


```{r}
weather_df |>
  ggplot(aes(x = prcp)) +
  geom_histogram()
```


```{r}
weather_df |>
  filter(prcp > 1000)
```

```{r}
weather_df |>
  filter(tmax > 20, tmax < 30) |>
  ggplot(aes(x = tmin, y = tmax, color = name, shape = name)) +
  geom_point()
```

Strange: There are bars between every value in Central Park and Molokai

Reason: The way of reporting data in Waterhole is different with Central Park and Molokai

## group_by()
counting stuff

```{r}
weather_df |>
  group_by(name) |>
  summarize(
    n_obs = n(), 
    n_dist = n_distinct(month)
    )
```

## 2x2 Table

```{r}
weather_df |>
  drop_na(tmax) |>
  mutate(
    cold = case_when(
      tmax < 5 ~"cold",
      tmax >= 5 ~ "not cold"
    )
  ) |>
  group_by(name, cold) |>
  summarize(count = n())

weather_df |>
  filter(name != "Molokai_HI") |>
  mutate(
    cold = case_when(
      tmax < 5 ~"cold",
      tmax >= 5 ~ "not cold"
    )
  ) |>
  janitor::tabyl(name, cold)
```

```{r}
weather_df |>
  group_by(name, month) |>
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE),
    sd_prcp = sd(prcp, na.rm = TRUE)
  ) |>
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point() +
  geom_line()
```

Format for readers

```{r}
weather_df |>
  group_by(name, month) |>
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE)
  ) |>
  pivot_wider(
    names_from = name,
    values_from = mean_tmax
  ) |>
  knitr::kable(digits = 3)
```

## Grouped mutates

```{r}
weather_df |>
  group_by(name) |>
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    ## group level standardization
    centered_tmax = tmax - mean_tmax
  ) |>
  ggplot(aes(x = date, y = centered_tmax, color = name)) +
  geom_point()
```

### Find hottest/coldest days

```{r}
weather_df |>
  mutate(
    ## rank from the smallest number
    temp_rank = min_rank(tmax)
    ## hotest days
    ## temp_rank = min_rank(desc(tmax))
  ) |>
  filter(temp_rank < 10)

weather_df |>
  group_by(name) |>
  filter(min_rank(tmax) < 4) |>
  arrange(tmax)
```

lag()

```{r}
weather_df |>
  group_by(name) |>
  mutate(
    lagged_tmax = lag(tmax),
    temp_change = tmax - lagged_tmax
  ) |>
  filter(min_rank(temp_change) < 3)

weather_df |>
  group_by(name) |>
  mutate(
    lagged_tmax = lag(tmax),
    temp_change = tmax - lagged_tmax
  ) |>
  summarize(
    sd_tmax_change = sd(temp_change, na.rm = TRUE)
  )
```

## LC

```{r}
pulse_df = read_sas("data/public_pulse_data.sas7bdat") |>
  janitor::clean_names() |>
  pivot_longer(
    cols = bdi_score_bl : bdi_score_12m,
    names_to = "visit",
    values_to = "bdi_score",
    names_prefix = "bdi_score_"
  )
  
pulse_df |>
  group_by(visit) |>
  summarize(
    mean_bdi = mean(bdi_score, na.rm = TRUE)
  ) |>
  knitr::kable(digits = 1)

```


```{r}
litters_df = read_csv("data/FAS_litters.csv", na = c("NA", "", ".")) |>
  janitor::clean_names() |>
  separate(
    group, into = c("dose", "tx_day"), sep = 3
  )

pups_df = read_csv("data/FAS_pups.csv", na = c("NA", "", ".")) |>
  janitor::clean_names()

fas_df = left_join(pups_df, litters_df, by = "litter_number")
```

Compute table:
for now, focus on post-natal day on which a pup is able to pivot. Two predictors of interest are the dose level and the day of treatment. Produce a reader-friendly table that quantifies the possible associations between dose, day of treatment, and the ability to pivot.

```{r}
fas_df |>
  drop_na(dose) |>
  group_by(dose, tx_day) |>
  summarize(mean_pivot = mean(pd_pivot, na.rm = TRUE)) |>
  pivot_wider(
    names_from = tx_day,
    values_from = mean_pivot
  ) |>
  knitr::kable(digits = 2)
```










